---
title: "Peninsular Spanish Audios Analysis"
author: "Kaylee Fernandez"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
suppressPackageStartupMessages({ library(mgcv); library(lme4) })
input_path <- "/Users/kayleefernandez/Spanish_F0_study_audios/praat_output_3_words.txt"
source("data_analysis.R")

# Data
n_part  <- nlevels(factor(praat_clean$participant))
n_items <- dplyr::n_distinct(praat_clean$item_id)
n_ns2   <- nrow(ns2_tok)
alpha       <- .05
bonf_alpha  <- .05 / 4
struct_lab <- function(s) gsub("\\|", "\u2502", s)
gamm_mode  <- if (isTRUE(f0_fit$pilot)) "descriptive (pilot: by-participant smooths dropped)" else "inferential"

# APA p formatter
apa_p <- function(p) {
  if (is.na(p)) return("= NA")
  if (p < .001) return("< .001")
  paste0("= ", sub("^0", "", formatC(p, format = "f", digits = 3)))
}

# Pull GAMM parametric (height) and smooth (shape) terms by name
gamm_terms <- function(fit) {
  sm <- summary(fit$model)
  pr <- sm$p.table; st <- sm$s.table
  h  <- grep("conditionO", rownames(pr))
  sh <- grep("conditionO", rownames(st)); sh <- sh[!grepl("participant", rownames(st)[sh])][1]
  list(
    height_b = pr[h, "Estimate"], height_t = pr[h, 3], height_p = pr[h, 4],
    shape_F = st[sh, "F"], shape_edf = st[sh, "edf"], shape_rdf = st[sh, "Ref.df"],
    shape_p = st[sh, "p-value"])
}

# Pull GLMM condition effect (estimate from summary, chi-square + p from NMC)
glmm_terms <- function(pair) {
  co <- summary(pair$inclusive)$coefficients
  b  <- co[grep("^conditionb$", rownames(co)), "Estimate"]
  a  <- anova(pair$restricted, pair$inclusive)
  list(b = b, chisq = a$Chisq[2], df = a$Df[2], p = a[["Pr(>Chisq)"]][2])
}

f0t  <- gamm_terms(f0_fit)
intt <- gamm_terms(int_fit)
durG <- glmm_terms(dur_pair)
intG <- glmm_terms(int_pair)
f0G  <- glmm_terms(f0_pair)
shp  <- gamm_terms(list(model = m_f0_shape))

# Significance-gated sentence builder: emit a finding only if p < alpha
sig_or_null <- function(p, sig_text, null_text, thr = alpha) {
  if (!is.na(p) && p < thr) sig_text else null_text
}

dur_struct <- struct_lab(dur_pair$structure)
int_struct <- struct_lab(int_pair$structure)
f0_struct  <- struct_lab(f0_pair$structure)
```

# Design

`r n_part` participants, `r n_items` two-syllable-noun items, each read in both conditions and in both lists. Condition `a` = present/paroxytone, `b` = past/oxytone. Everyone read list L1 first, so list = presentation order (entered centered as `order`). Effect direction is past − present. Decision rule: alpha = .05; for the four N-s2 cue tests, Bonferroni alpha = `r sub("^0","",formatC(bonf_alpha, format="f", digits=4))`. Contour GAMMs were `r gamm_mode`. Only effects reaching significance are reported below; non-significant tests are noted as null.

# Contour models (F0, intensity)

**Type.** Non-parametric (GAMM; `mgcv::bam`, fREML, AR(1) error correction); a parametric height term plus a non-parametric difference smooth.
**DV.** F0 (semitones re 100 Hz); intensity (dB).
**IVs (fixed).** `conditionO` (ordered) carrying the height term and the `s(time, by = conditionO)` shape smooth; `order` (covariate).
**Random intercepts.** by item. **Random slopes.** by-participant reference and difference factor smooths.

## F0

```{r f0-text, results='asis'}
parts <- c()
if (f0t$height_p < alpha) {
  parts <- c(parts, sprintf("a present/past difference in contour height (b = %.2f st, *t* = %.2f, *p* %s)",
                            f0t$height_b, f0t$height_t, apa_p(f0t$height_p)))
}
if (f0t$shape_p < alpha) {
  parts <- c(parts, sprintf("a difference in contour shape, s(time):conditionO (*F*(%.2f, %.2f) = %.2f, *p* %s)",
                            f0t$shape_edf, f0t$shape_rdf, f0t$shape_F, apa_p(f0t$shape_p)))
}
if (length(parts)) {
  cat(sprintf("There was %s. The difference was localized to the verb.\n", paste(parts, collapse = " and ")))
} else {
  cat("No present/past F0 difference reached significance.\n")
}
```

```{r f0-fig, fig.cap="F0 contour (A) and present−past difference (B)."}
print(fig_f0)
```

## Intensity

```{r int-text, results='asis'}
parts <- c()
if (intt$height_p < alpha) {
  parts <- c(parts, sprintf("height (b = %.2f dB, *t* = %.2f, *p* %s)",
                            intt$height_b, intt$height_t, apa_p(intt$height_p)))
}
if (intt$shape_p < alpha) {
  parts <- c(parts, sprintf("shape, s(time):conditionO (*F*(%.2f, %.2f) = %.2f, *p* %s)",
                            intt$shape_edf, intt$shape_rdf, intt$shape_F, apa_p(intt$shape_p)))
}
if (length(parts)) {
  cat(sprintf("There was a present/past difference in %s, localized to the verb.\n", paste(parts, collapse = " and ")))
} else {
  cat("No present/past intensity difference reached significance.\n")
}
```

```{r intensity-fig, fig.cap="Intensity contour (A) and present−past difference (B). Red points mark the region where the 95% CI excludes zero."}
print(fig_int)
```

# N-s2 anticipation models (duration, intensity, F0 mean)

**Type.** Parametric (LMM; `lme4::lmer`, ML). Each condition effect assessed by nested model comparison, `anova(restricted, inclusive)` (likelihood-ratio test), reporting df, chi-square, *p*.
**DV.** N-s2 duration (ms); mean intensity (dB); mean F0 (st).
**IVs (fixed).** `condition`; `order` (covariate).
**Random structure.** Maximal `(1 + condition | participant) + (1 + condition | item_id)` attempted, stepping down when singular; structure used is reported per model. Based on `r n_part` participants, `r n_ns2` N-s2 tokens.

## Duration

```{r dur-text, results='asis'}
cat(sprintf("Random structure used: %s. ", dur_struct))
surv <- if (durG$p < bonf_alpha) ", and remains significant after Bonferroni correction" else " (not significant after Bonferroni correction)"
cat(sig_or_null(durG$p,
  sprintf("N-s2 was shorter before an oxytone verb (b = %.2f ms; NMC: *\u03c7\u00b2*(%d) = %.2f, *p* %s)%s.",
          durG$b, durG$df, durG$chisq, apa_p(durG$p), surv),
  sprintf("No condition difference in N-s2 duration (b = %.2f ms; NMC: *\u03c7\u00b2*(%d) = %.2f, *p* %s).",
          durG$b, durG$df, durG$chisq, apa_p(durG$p))))
```

```{r dur-fig, fig.cap="N-s2 duration by condition: each participant's means (gray) and the population-level pooled estimate (black)."}
print(p_dur)
```

## Intensity

```{r ns2int-text, results='asis'}
cat(sprintf("Random structure used: %s. ", int_struct))
cat(sig_or_null(intG$p,
  sprintf("Condition effect on N-s2 intensity (b = %.2f dB; NMC: *\u03c7\u00b2*(%d) = %.2f, *p* %s).",
          intG$b, intG$df, intG$chisq, apa_p(intG$p)),
  sprintf("No condition difference in N-s2 intensity (b = %.2f dB; NMC: *\u03c7\u00b2*(%d) = %.2f, *p* %s).",
          intG$b, intG$df, intG$chisq, apa_p(intG$p))))
```

## F0 mean (height)

```{r ns2f0-text, results='asis'}
cat(sprintf("Random structure used: %s. ", f0_struct))
cat(sig_or_null(f0G$p,
  sprintf("Condition effect on mean N-s2 F0 (b = %.2f st; NMC: *\u03c7\u00b2*(%d) = %.2f, *p* %s).",
          f0G$b, f0G$df, f0G$chisq, apa_p(f0G$p)),
  sprintf("No condition difference in mean N-s2 F0 (b = %.2f st; NMC: *\u03c7\u00b2*(%d) = %.2f, *p* %s).",
          f0G$b, f0G$df, f0G$chisq, apa_p(f0G$p))))
```

# N-s2 F0 contour shape

**Type.** Non-parametric (GAMM; `mgcv::bam`, fREML). Tests whether the F0 trajectory *through* N-s2 differs by condition.
**DV.** F0 (st), voiced N-s2 points.
**IVs (fixed).** `conditionO` + `s(t)` + `s(t, by = conditionO)`; `order` (covariate); `t` = within-syllable position.
**Random intercepts.** by item. **Random slopes.** by-participant reference and difference factor smooths.

```{r shape-text, results='asis'}
if (shp$shape_p < alpha) {
  surv <- if (shp$shape_p < bonf_alpha) "and remains significant after Bonferroni correction" else "but is not significant after Bonferroni correction, so it is reported as suggestive only"
  cat(sprintf("The within-syllable F0 tilt differed by condition, s(t):conditionO (*F*(%.2f, %.2f) = %.2f, *p* %s), %s.\n",
              shp$shape_edf, shp$shape_rdf, shp$shape_F, apa_p(shp$shape_p), surv))
} else {
  cat(sprintf("No within-syllable F0 tilt difference reached significance (*F*(%.2f, %.2f) = %.2f, *p* %s).\n",
              shp$shape_edf, shp$shape_rdf, shp$shape_F, apa_p(shp$shape_p)))
}
```

```{r shape-fig, fig.cap="Predicted N-s2 F0 contour by condition."}
print(p_f0shape)
```

# Summary of significant findings

```{r summary-table, results='asis'}
rows <- list()
add <- function(region, type, dv, effect, test, p) {
  rows[[length(rows) + 1]] <<- c(region, type, dv, effect, test, apa_p(p))
}
if (f0t$height_p < alpha) add("Verb F0 height","GAMM","F0 (st)", sprintf("%.2f", f0t$height_b), sprintf("t = %.2f", f0t$height_t), f0t$height_p)
if (f0t$shape_p  < alpha) add("Verb F0 shape","GAMM","F0 (st)","tilt/shape", sprintf("F = %.2f", f0t$shape_F), f0t$shape_p)
if (intt$height_p < alpha) add("Verb intensity height","GAMM","Int (dB)", sprintf("%.2f", intt$height_b), sprintf("t = %.2f", intt$height_t), intt$height_p)
if (intt$shape_p  < alpha) add("Verb intensity shape","GAMM","Int (dB)","tilt/shape", sprintf("F = %.2f", intt$shape_F), intt$shape_p)
if (durG$p < alpha) add("N-s2 duration","LMM","Dur (ms)", sprintf("%.2f", durG$b), sprintf("\u03c7\u00b2(%d) = %.2f", durG$df, durG$chisq), durG$p)
if (intG$p < alpha) add("N-s2 intensity","LMM","Int (dB)", sprintf("%.2f", intG$b), sprintf("\u03c7\u00b2(%d) = %.2f", intG$df, intG$chisq), intG$p)
if (f0G$p  < alpha) add("N-s2 F0 mean","LMM","F0 (st)", sprintf("%.2f", f0G$b), sprintf("\u03c7\u00b2(%d) = %.2f", f0G$df, f0G$chisq), f0G$p)
if (shp$shape_p < alpha) add("N-s2 F0 shape","GAMM","F0 (st)","tilt", sprintf("F = %.2f", shp$shape_F), shp$shape_p)
if (length(rows)) {
  tbl <- as.data.frame(do.call(rbind, rows), stringsAsFactors = FALSE)
  names(tbl) <- c("Region / cue","Type","DV","Effect (b-a)","Test","p")
  cat(knitr::kable(tbl, format = "pipe"), sep = "\n")
} else {
  cat("_No effects reached significance._\n")
}
```

```{r summary-prose, results='asis'}
verb_sig <- (f0t$height_p < alpha | f0t$shape_p < alpha | intt$height_p < alpha | intt$shape_p < alpha)
dur_sig  <- durG$p < alpha
cat("\n\n")
if (verb_sig) cat("The present/past stress contrast surfaces acoustically at the verb. ")
if (dur_sig)  cat("The noun's final syllable carries a durational anticipatory cue (shorter before oxytones). ")
cat("Non-significant cues are reported as null above.\n")
```
